---
title: "Identyfing all Important predictor for Heart Disease"
output:
editor_options: 
  markdown: 
    wrap: 72
---

| Aim :                                                                              |
|:-------|
| To classify which all attributes are the most important predictor of heart disease |

## Dataset Attributes

Age : age of the patient [years] Sex : sex of the patient [M: Male, F:
Female] ChestPainType : chest pain type [TA: Typical Angina, ATA:
Atypical Angina, NAP: Non-Anginal Pain, ASY: Asymptomatic] RestingBP :
resting blood pressure [mm Hg] Cholesterol : serum cholesterol [mm/dl]
FastingBS : fasting blood sugar [1: if FastingBS \> 120 mg/dl, 0:
otherwise] RestingECG : resting electrocardiogram results [Normal:
Normal, ST: having ST-T wave abnormality (T wave inversions and/or ST
elevation or depression of \> 0.05 mV), LVH: showing probable or
definite left ventricular hypertrophy by Estes' criteria] MaxHR :
maximum heart rate achieved [Numeric value between 60 and 202]
ExerciseAngina : exercise-induced angina [Y: Yes, N: No] Oldpeak :
oldpeak = ST [Numeric value measured in depression] ST_Slope : the slope
of the peak exercise ST segment [Up: upsloping, Flat: flat, Down:
downsloping] HeartDisease : output class [1: heart disease, 0: Normal]

Importing all the necessary libraries:

```{r}
library(tidyverse)
library(lubridate)
library(nycflights13)
library(ggplot2)
library(dplyr)
library(plotly)
library(readr)
```

Data Info:

```{r}
heart <- read_csv("C:/Users/ROSHAN D K/Desktop/R project/heart.csv", show_col_types = FALSE)
heart
```

Checking the dimension of the dataset

```{r}
dim(heart)
```

Finding the data type of each **column**

```{r}
str(heart)
```

Checking Missing values in each column

```{r}
heart %>%
  summarise_all(~ sum(is.na(.)))
```

Well,There is no null values found in the dataset

Now, We are going to divide the data into two features Categorical
Features : Sex ChestPainType FastingBS RestingECG ExerciseAngina
ST_Slope, HeartDisease

```{r}
# Create a data frame with only numerical columns
heart_num <- select_if(heart, is.numeric)

# Create a data frame with only categorical columns
heart_cat <- heart[, sapply(heart, function(x) !is.numeric(x))]

```

Numerical Features : Age, RestingBP, Cholesterol, MaxHR, Oldpeak. If an
attribute contains fewer than six distinct elements, it is characterised
as a categorical feature; otherwise, it is a numerical feature.

# Exploratory Data Analysis

Exploring Overall Distribution of data

```{r}
# Create a bar plot to visualize the percentages
library(scales)
heart_disease_percent <- prop.table(table(heart$HeartDisease))
bp <- ggplot(heart, aes(x="", y=HeartDisease, fill=heart$Sex)) +
geom_bar(width = 1, stat = "identity")
pie <- bp + coord_polar("y", start=0)
pie + scale_fill_manual(values=c("#E69F00", "#56B4E9"))
  geom_text(aes(y = HeartDisease/3 + c(0, cumsum(HeartDisease)[-length(HeartDisease)]), label = HeartDisease), size=5)
  
# Create a frequency table for the output class
table(heart$HeartDisease)
# Create a frequency table for the 'Sex' variable
table(heart$Sex, heart$HeartDisease)

```

This dataset contains more number of Males observation than the female,
so we need to keep this in mind. observations of male with heart disease
is 8 times more than a female. people with heart disease and people with
no heart disease is preety much evenly distributed.

# Categorical Feature Analysis

Target variable visualisation All the observations contains after the
FastingBS output.

```{r}
ggplot(heart, aes(x=Sex, fill=factor(HeartDisease))) +
  geom_bar(position = "dodge") +
  ggtitle("Sex vs Heart Disease")

ggplot(heart, aes(x=ChestPainType, fill=factor(HeartDisease))) +
  geom_bar(position = "dodge") +
  ggtitle("ChestPainType vs Heart Disease")

ggplot(heart, aes(x=RestingECG, fill=factor(HeartDisease))) +
  geom_bar(position = "dodge") +
  ggtitle("RestingECG vs Heart Disease")

ggplot(heart, aes(x=ExerciseAngina, fill=factor(HeartDisease))) +
  geom_bar(position = "dodge") +
  ggtitle("ExerciseAngina vs Heart Disease")

ggplot(heart, aes(x=ST_Slope, fill=factor(HeartDisease))) +
  geom_bar(position = "dodge") +
  ggtitle("ST_Slope vs Heart Disease")

ggplot(heart, aes(x=FastingBS, fill=factor(HeartDisease))) +
  geom_bar(position = "dodge") +
  ggtitle("FastingBS vs Heart Disease")

```

Observations: 1. Men are nearly two times as likely as women to suffer
heart disease.

2.  There are about 500 cases of asymptomatic chest discomfort, which
    means that half of our participants had no prior symptoms before
    having a heart attack.

3.  Resting electrocardiogram: Person with ST-T wave abnormalities is
    more likely to have a cardiac condition than the others.

4.  ExerciseAngina: Compared to exercise-induced angina with a "No,"
    exercise-induced angina with a "Yes" is about 2.4 times more likely
    to be associated with a cardiac condition.

5.  ST Slope: The peak workout ST segment's slope varies. Compared to
    the other two segments, ST Slope Up considerably less likely to have
    heart disease.

6.  FastingBS: Patients with and without fasting blood sugar diagnoses
    are much more likely to develop heart disease.

# Numerical Features Analysis:

All the ranges are given after all the visualizations Detecting Outliers

```{r}
# Box plot for the 'Age' variable
ggplot(heart, aes(x=factor(HeartDisease), y=Age)) +
  geom_boxplot() +
  ggtitle("Age vs Heart Disease")

# Box plot for the 'Age' variable
ggplot(heart, aes(x=factor(HeartDisease), y=RestingBP)) +
  geom_boxplot() +
  ggtitle("RestingBP vs Heart Disease")

# Box plot for the 'Age' variable
ggplot(heart, aes(x=factor(HeartDisease), y=Cholesterol)) +
  geom_boxplot() +
  ggtitle("Cholesterol vs Heart Disease")

# Box plot for the 'Age' variable
ggplot(heart, aes(x=factor(HeartDisease), y=MaxHR)) +
  geom_boxplot() +
  ggtitle("MaxHR vs Heart Disease")

# Box plot for the 'Age' variable
ggplot(heart, aes(x=factor(HeartDisease), y=Oldpeak)) +
  geom_boxplot() +
  ggtitle("Oldpeak vs Heart Disease")
```

Observations: The range at which patient have high chance of positive
cases of heart disease Age : 50+ RestingBP : 95 - 170 Cholesterol : 0 -
600 MaxHR : 70 - 180 Oldpeak : 0 - 4

Checking Impact of numerical feature on output

```{r}

ggplot(data = heart, aes(x= Age)) +
  stat_count(color = "darkblue", fill = "lightblue") +
  labs(title = "Age Distribution", x = "Age", y = "Count") +
  scale_x_log10()

ggplot(data = heart, aes(x= RestingBP)) +
  stat_count(color = "darkblue", fill = "lightblue") +
  labs(title = "RestingBP Histogram Plot", x = "Age", y = "Count")  +
  scale_x_log10()

ggplot(data = heart, aes(x= Cholesterol)) +
  stat_count(color = "darkblue", fill = "lightblue") +
  labs(title = "Cholesterol Histogram Plot", x = "Age", y = "Count")   +
  scale_x_log10()

ggplot(data = heart, aes(x= MaxHR)) +
  stat_count(color = "darkblue", fill = "lightblue") +
  labs(title = "MaxHR Histogram Plot", x = "Age", y = "Count")  +
  scale_x_log10()

ggplot(data = heart, aes(x= Oldpeak)) +
  stat_count(color = "darkblue", fill = "lightblue") +
  labs(title = "Oldpeak Histogram Plot", x = "Age", y = "Sex") 
```

The maximum density is found in people aged 50 to 65, with a mean age of
53.5. Oldpeak's data distribution is rightly skewed appropriately. The
data distribution for cholestrol is bidmodal.

# Bi-variate Analysis

```{r}
ggplot(heart, aes(x=Age, y=HeartDisease, color=Sex)) +
  geom_point() +
  geom_smooth(method = lm, se = FALSE) +
  labs(title="Age vs Heart disease")

```

It shows a clear trend that, as patients get older, they have more cases
of heart diseases.

```{r}
ggplot(heart, aes(x=MaxHR, y=HeartDisease, color=Sex)) +
  geom_point() +
  geom_smooth(method = lm, se = FALSE) +
  labs(title="Maximum Heart Rate Vs Heart Disease")

```

As expected, those with cardiac disease had lower maximum heart rates.
This might be due to the diseases themselves, which could lead to heart
degeneration.

```{r}
ggplot(heart, aes(x=Age, y=MaxHR, color=Sex)) +
  geom_point() +
  geom_smooth(method = lm, se = FALSE) +
  labs(title = "Age vs Maximum Heart Rate")

```

This is not directly associated with the heart disease but still good to
know: People's maximum heart rates normally decline with age since their
hearts aren't as strong as they were in their youth.

```{r}
ggplot(heart, aes(x=Age, y=RestingBP, color=Sex)) +
  geom_point() +
  geom_smooth(method = lm, se = FALSE) +
  labs(title = "Age vs Resting blood sugar level")
```

As patients get older their blood sugar level also increases

```{r}
ggplot(heart, aes(x=Cholesterol, y=HeartDisease, 
                                    color=Sex)) +
  geom_point() +
  geom_smooth(method = lm, se = FALSE) +
  labs(title="Cholesterol Levels Vs Heart Diseases")

```

Patients with heart disease had normal cholesterol levels. This is
uncommon because heart failure is often brought on by high cholesterol
levels.

```{r}
ggplot(heart, aes(x=FastingBS, y=HeartDisease, color=Sex)) +
  geom_point() +
  geom_smooth(method = lm, se = FALSE) +
  labs(title = "Sugar Blood Level Vs Heart Diseases")
```

As expected, Patients with heart diseases usually have a higher levels
of blood sugar sugar levels.

```{r}
ggplot(heart, aes(x=Oldpeak, y=HeartDisease, color=Sex)) +
  geom_point() +
  geom_smooth(method = lm, se = FALSE) +
  labs(title = "Oldpeak vs HeartDisease")
```

Trend is very clear that, patients with heart disease have higher
oldpeak (Its a Numeric value measured in depression)

```{r}
ggplot(heart, aes(x=RestingBP, y=HeartDisease, color=Sex)) +
  geom_point() +
  geom_smooth(method = lm, se = FALSE) +
  labs(title = "RestingBP vs HeartDisease")
```

Patients with higher blood pressure level has higher cases of having
heart disease.

Multivariate Analysis

```{r}
# Scatter plot for the 'RestingBP' and 'Cholesterol' variables
ggplot(heart, aes(x=FastingBS, y=Cholesterol, color=factor(HeartDisease))) +
  geom_point() +
  geom_smooth(method = lm, se = FALSE) +
  ggtitle("Fasting blood sugar level vs Cholesterol vs Heart Disease")

ggplot(heart, aes(x=Cholesterol, y=Oldpeak, color=factor(HeartDisease))) +
  geom_point() +
  geom_smooth(method = lm, se = FALSE) +
  ggtitle("Cholesterol vs Resting Blood Pressure vs Heart Disease")

ggplot(heart, aes(x=RestingBP, y=MaxHR, color=factor(HeartDisease))) +
  geom_point() +
  geom_smooth(method = lm, se = FALSE) +
  ggtitle("Resting Blood Pressure vs Maximum Heart Rate vs Heart Disease")

ggplot(heart, aes(x=Age, y=RestingBP, color=factor(HeartDisease))) +
  geom_point() +
  geom_smooth(method = lm, se = FALSE) +
  ggtitle("Age vs Resting Blood Pressure vs Heart Disease")

ggplot(heart, aes(x=Age, y=Cholesterol, color=factor(HeartDisease))) +
  geom_point() +
  geom_smooth(method = lm, se = FALSE) +
  ggtitle("Age vs Cholesterol level vs Heart Disease")


```

1.  Patients are more likely to develop heart disease if their maximum
    heart rate is higher and their resting blood pressure is lower.
2.  As we already know, as people age, their blood sugar levels rise as
    well, which also tends to raise their risk of developing heart
    disease.

```{r}
# # Fit a decision tree model using all of the variables   
library(rpart)
library(rpart.plot)
model <- rpart(HeartDisease ~ ., data = heart)
rpart.plot(model)
```

Statistical testing

The chi-square test to determine if there is a significant association
between all the numerical feature and HeartDisease

```{r}
# Chi-square test
library(vcd)
chisq.test(table(heart$Cholesterol, heart$HeartDisease))
chisq.test(table(heart$MaxHR, heart$HeartDisease))
chisq.test(table(heart$FastingBS, heart$HeartDisease))
chisq.test(table(heart$Age, heart$HeartDisease))
```

Observations: As p-value less than 0.05 would indicate a significant
association between the heart disease and all thease features like
Cholestrol level, Maximum heart rate, Age and finally blood sugar level.

For some reason I am not able to knit it using this Chunck so I am
commenting it out, when you remove the comment model is working fine.

```{r}
# Split the data into training and test sets
#set.seed(123)
# split_index <- createDataPartition(heart$HeartDisease, p = 0.7, list = FALSE)
#train_data <- heart[split_index, ]
#test_data <- heart[-split_index, ]

# Fit a logistic regression model
#model <- glm(HeartDisease ~ Age + Sex + ChestPainType + RestingBP + Cholesterol + FastingBS + RestingECG + MaxHR + ExerciseAngina + Oldpeak + ST_Slope, data = train_data, family = binomial())

# Summary of the model
#summary(model)
```

Observations: Age, RestingBP, FastingBS, ExerciseAngina (Yes),Oldpeak
and ST_SlopeFlat feature positively correlate with the heart disease,
whereas other feature negatively correlate with the heart disease.

```{r}
```

Conclusion:

Out of all heart disease patients, a startling 90% of them are men,
according to the data. 77% of ASY types result in cardiac problems,
which is the majority. Exercise-induced angina with a 'Yes' is about 2.4
times more likely to suffer a heart disease. A fasting blood sugar level
below 120 mg/dl indicates a greater risk of developing heart disease.
From the perspective of numerical characteristics, patients in this
range are more likely to suffer heart illness. Age: 50+, Resting Blood
Pressure: 95-170, Cholesterol: 0--600, Maximum Heart Rate: 70--180,
Oldpeak: 0--4. Age, Maximum Heart Rate, Cholesterol, and Fasting Blood
Sugar Level are the most significant predictors of heart disease,
according to our analysis of various features

The goal of researching the association between these characteristics
and heart disease is to get a better knowledge of the potential risk
poses by this feature on heart disease and to build reliable models for
predicting heart disease.There were a few difficulties with this
analysis, including: 1. Getting a big, diverse dataset for analysis that
appropriately represents the community of interest is one of the biggest
obstacles. 2. The quality of the data gathered is a significant
difficulty; in our instance, the study of the cholestrol variable,
inaccurate data might have a substantial impact on the findings. 3.
Another issue in this research is selecting the most relevant factors,
as the number of variables might be fairly vast, and not all variables
may be useful for the categorization of heart disease. 4. I've simply
used the logistics regression model in my example, but choosing the best
model for this study is difficult because it can have a big impact on
the outcomes. 5. Finally, applying domain knowledge from the medical
field to the outcomes.

Few Mistake that were made during the analysis, was not properly dealing
with the outlier before deploying the model like applying dimensionality
reduction theorem on the outliers or any other technique. If I had more
time to complete the research, I would have conducted a more thorough
analysis of each variable's distribution, as well as a deeper
investigation into the association between each variable and heart
disease using a random forest model and other sophisticated statistical
techniques.

Another research subject I'd want to look at is: To determine the
relationship between different variables and the likelihood of heart
disease. Using a machine learning technique, to discover the most
important predictors of heart disease. To investigate the effect of
demographic parameters on the risk of heart disease, such as age and
gender. Using predictive modelling tools, uncover possible early warning
indicators for heart disease.

The same research question(s) could have been addressed differently if
data contain 1. The smoking habbits of the patients. 2. Also, contains
the variables like patients were on any medication or had any other
disease 3. Data would give more insight if it contains the occupation
and more demographic data. 4. data on Eating habbits could also be
usefull. 5. Data about if they drink alchol or not.

References
<https://www.kaggle.com/code/durgancegaur/a-guide-to-any-classification-problem>
<https://www.kaggle.com/code/tanmay111999/heart-failure-prediction-cv-score-90-5-models>
<https://www.kaggle.com/code/azizozmen/heart-failure-predict-8-classification-techniques>
<https://www.kaggle.com/code/pythonafroz/evaluation-metrics-used-in-machine-learning>
